---
title: "A Population-based study of Hospital Readmission Risk for American patients
"
date: "December,20, 2020"
auther: Yuanzhe Yang
output: html_document

---
## Introduction

People’s daily lifestyles have been enormously improved since the early 21st century, due to the rapid development of global economy and scientific techniques. However, people are suffering the curse of advanced technology at the same time: improperly managed diet, insufficient physical excursive and other unhealthy behaviors bring people in a risk of common disease: diabetes. Although this chronic disease does not present apparent symptoms, it has no specific medical treatment for instant cure at present and diabetic patients tend to be more sensitive to be involved in other dangerous diseases, for instance, heart disease or pneumonia, due to their vulnerable immune system.  In the United States, diabetic patients averagely cost the Medicare extra 250 million dollars spend for their readmitted retreatment since 2011 [1], which definitely increased a financial burden on federal government. On the other hand, the rate of readmitted patients is commonly utilized as a criterion of evaluating the performance of a specific hospital: the poor diagnosis or invalid treatment could be distinguished by the high readmission rate of patients [2]. Consequently, investigate the relationship between hospital readmission and diabetes, and evaluate the influence of potential factors related with the risk of readmission could provide some profound insight for reducing the federal cost while also providing sufficient support to patients with more reasonable arrangement of medical resource. 

Generally, several methods and programs had been utilized for controlling repeated hospitalization in recent years. Unplanned readmission risk could be decreased by more tightened discharge rules as proposed by Goudjerkan et al. [3]. The U.S. Centers for Medicare and Medicaid Services (CMS) has conducted an initiative program called Hospital Readmission Reeducation Program (HRRP) since 2012, sufficiently decreased the rate of repeated encounters [4]. 

In this report, we initiated with our study with data cleaning and preliminary analysis on the dataset from the UCI Machine Learning Repository. The response variable we taken was “Unplanned hospital readmission”, in general, if a patient had a readmission in less than 30 days, it would be identified as a case of unplanned hospital readmission. Then we constructed models by Generalized Linear Regression (GLM) and Generalized linear mixed model (GLMM), respectively, for evaluating the potential factors that may be related to hospital readmission.  The regression results showed that and we concluded that the patients’ medical history, clinical care records and the factor of diabetes were statistically significant related with the risk of readmission.


## Method
### Data

We focused on a data set from the UCI Machine Learning Repository (Data source:
https://archive.ics.uci.edu/ml/datasets/diabetes). The dataset originally contains national
clinical information data collected from over 130 US hospitals in the period from 1999 to 2008.
71,518 patients had been traced with single or multiple records, 101,766 observations were
included with 49 features, such as demographic and geographic information, history medical
records, whether a patient was readmitted or not, etc. The index for hospital admission, which
was our main interest, was assessed by the number of encounter in this dataset. And we did
notice that 16,733 surveyed patients, that is, approximately 23.40% of them eventually had
one or multiple experiences of readmission. For these readmitted US patients, we also
observed that approximately 62.21% of them were commonly readmitted by twice.

#### Data cleaning

The process of data cleaning is inevitably necessary for our original dataset because a
substantial proportion of missing values has been found in several covariates as shown in
Figure 2. We directly removed the variables weight, prayer_code and medical_speciality since
they seem to contain more than 20% unregistered records, even we only randomly sampled
2,000 observations from the pool. Also, the missing values contained in the variable of race
are removed. Furthermore, since our primary target of this study was exploring the
determinants that may associated with the risk of “Unplanned hospital readmission”, in
general, if a patient had a readmission in less than 30 days, it would be identified as a case of
unplanned hospital readmission. Then we created a new binary variable as follows:

The cleaning and reformatted dataset eventually contained 69,668 patients with 99,493
observations and 46 variables. For exploring the validation and accuracy of our GLM and
GLMM methods in later process, we randomly divided our dataset by a common ratio 80/20
as training and testing datasets, respectively.

```{r}
setwd("~/Desktop/Data")

# package used for analysis
library(tidyverse)
library(ggplot2)
library(visdat)
library(glmnet)
library(foreach)
library(lme4)
library(pROC)
library(ggpubr)
library(lattice)


# read data into R
df1 <- read.csv("diabetes_update.csv")
# brief view
View(df1)
head(df1)
# dimension of dataframe
dim(df1)
# number of patients 
length(table(df1$patient_nbr))
# attribution of dataframe
str(df1)
# check NA values in dataframe
sapply(df1, function(x) sum(is.na(x)))

# proportion of NA values

sum(is.na(df1$race))/nrow(df1)

sum(is.na(df1$weight))/nrow(df1)

sum(is.na(df1$payer_code))/nrow(df1)

sum(is.na(df1$medical_specialty))/nrow(df1)

# distribution of number of encounters 
counten <-df1 %>%
  group_by(patient_nbr) %>%
  filter(n() > 0) %>%
  count()

table(counten$n)
sum(table(counten$n)) - 54745
(sum(table(counten$n)) - 54745)/sum(table(counten$n))
10434/16773

# or we can compute by the number of encounter
table(df1$encounter_num)
length(table(df1$encounter_num))
a <- as.numeric(table(counten$n))
b <- a[-1]
p <- b/sum(b)
labs <- paste(round(p*100,2),"%",sep="")
labs
ns <- names(table(counten$n))
ns <- ns[-1]

# barplot of number of encounters
barplot(b, col = "slategray3", ylim = c(0, 12000), ylab = "Frequency",
        names.arg = ns)
table(counten$n)/sum(table(counten$n)) *100
text(0.8,11500, "62.21%", cex = 0.6)
text(2,4000, "19.84%", cex = 0.6)
text(3.2,2050, "8.47%", cex = 0.6)
text(4.4,1500, "4.27%", cex = 0.6)
text(5.6,1000, "2.06%", cex = 0.6)
text(6.8,800, "1.23%", cex = 0.6)

# checking missing value
id <- sample(1:nrow(df1), 2000, replace = FALSE)
df_na <- subset(df1, df1$X %in% id)
# remove missing values 
df2 <- subset(df1, select=-c(weight, payer_code, medical_specialty))
df3 <- df2[complete.cases(df2$race),]
# dimension of new dataset
dim(df3)

# indicator of readmission in 30 days
y <- rep(NA, nrow(df3))
for(i in 1 : nrow(df3)){
  if(df3$readmitted[i] == "<30"){
    y[i] = 1
  }
  else{
    y[i] = 0
  }
}

df.new <- df3
df.new$y <- ifelse(df3$readmitted == "<30", 1, 0)
df.new <- subset(df.new, select = -readmitted)
```


#### Preliminary analysis

Before initiating our modeling, we had a brief overview on our dataset. Table 1 summarized
the frequency and relative frequency of the unplanned hospital readmission on categorical
variables, such as gender and age groups, the implementation of diabetes medication. Besides,
mean and standard deviation were computed for continuous variables, for instance, the
inpatient days and number of lab procedures. From Table 1, we found that the odds ratio of
being readmitted for female was slightly greater than the one for male patients; the proportion
of readmission within 30 days did increase along with older patients. For the category of
ethnicity, no obvious pattern could be intuitively detected. Furthermore, the odds ratio of
readmission between diabetic and non-diabetic patients was greater than 1, which was
computed by (9.00/67.89)/(2.22/20.89), indicating a likelihood that the diabetic patients
would be readmitted to hospital at a higher risk than other patient. Meanwhile, we also noticed
that there were differences in the number of inpatients and emergency visits between
readmitted patients and non-readmitted patients, implying that the probability of
readmission might be infected by these factors. However, we should notice that the dataset
had an imbalanced issue since the readmission rate was only 11.22% out of the whole dataset,
therefore, the accuracy and efficiency of our later modelization need to be elaborately
examined.

```{r}
# descriptive statistics
table(y)
prop.table(table(y))
table(df3$gender, y)
prop.table(table(df3$gender, y))
table(df3$age, y)
prop.table(table(df3$age, y))
table(df3$race, y)
prop.table(table(df3$race, y))

table(df3$diabetesMed, y)
prop.table(table(df3$diabetesMed, y))
table(df3$change, y)
prop.table(table(df3$change, y))
aggregate(df3$num_lab_procedures, list(y), mean)
aggregate(df3$num_lab_procedures, list(y), sd)

aggregate(df3$Length.of.Sta, list(y), mean)
aggregate(df3$Length.of.Stay, list(y), sd)

aggregate(df3$num_medications, list(y), mean)
aggregate(df3$num_medications, list(y), sd)

aggregate(df3$number_inpatient, list(y), mean)
aggregate(df3$number_inpatient, list(y), sd)

aggregate(df3$number_emergency, list(y), mean)
aggregate(df3$number_emergency, list(y), sd)

prop.table(table(df3$gender, y),margin = 1)
a <- c(0.8869, 0.1131, 0.8888, 0.1112)


df3$readmitted <- y
df3 <- df3[-29794,]
df3$gender <- factor(df3$gender, levels = c("Female", "Male"))

# plots
# age group
p1 <- df3 %>% ggplot(mapping = aes(x = as.factor(age), fill = as.factor(readmitted))) +
    geom_bar() +
  labs(
       x = "Age",
       y = "Counts ") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), legend.position = "none")
  
# race group

p3 <- df3 %>% ggplot(mapping = aes(x = as.factor(race), fill = as.factor(readmitted))) +
    geom_bar() +
  labs(
       x = "Race",
       y = "Counts ") +
scale_fill_discrete(name = "Readmitted", labels = c("No", "Yes"))


# gender group
p2 <- df3 %>% ggplot(mapping = aes(x = as.factor(gender), fill = as.factor(readmitted))) +
    geom_bar() +
  labs(
       x = "Gender",
       y = "Counts ") +
  theme(legend.position = "none")

ggarrange(p1, p2, p3, nrow = 1, ncol = 3)
# bar plot 
yy <- y[-29794]
g <- df3$gender[-29794]
g <- factor(g, levels = c("Female", "Male"))
barplot(table(yy, g))

# create organized dataframe

df <-subset(df.new, select=-c(X, encounter_id, admission_source_id, encounter_num, examide,citoglipton))

df.new <-subset(df.new, select=-c(X, encounter_id, admission_source_id, encounter_num, examide,citoglipton))

df <- df[-29794,]

# training and testing dataset 
set.seed(333)
train.id = sample(1 : nrow(df), 0.8*nrow(df), replace = FALSE)
train = df[train.id, ]
test = df[-train.id, ]

```

## Models

### Variable selection

In order to allocate an appropriate combination of explanatory variables for fitting our model,
we apply the method of forward and backward stepwise regression proposed by Efroymson
[5]. The algorithms within this selection process is straightforward by a series of criteria, such
as AIC, BIC and Mallows's Cp. By applying R programming, we have a stepwise selected logistic
model with 9 variables: Number of inpatient visits, Diabetes medication prescribed, Number
of diagnoses, Number of diagnoses, Metformin, Discharge disposition, Length of stay, Number
of medication and Insulin.

#### Variance Component Model, GLM and GLMM

For detecting the variation in the tendency of being readmitted that could be attributed to the
individuals, for a binary response, it has a basic form as follows:

where $\pi_ij$ is the probability of 𝑖th individual being readmitted to hospital within 30 days after
𝑗th encounter. 𝜇" it the deviation in the likelihood of readmission for the 𝑖th individual from
the average. 𝜀"# is the term of error. If a random effect within each individual did exist, an
ANOVA test between a null model and a variance component model should have a significant
statistic to reject the null hypothesis that the two models have the same performance on fitting
data, then random effect across different patients on the likelihood of being readmitted could
be confirmed.

Compared to a common generalized linear model (GLM), generalized linear mixed model
(GLMM) is an extended form that included both fixed and random effects of covariates. Here,
we consider a GLMM containing the individual effect, that is, the average likelihood of
readmission was estimated by each random effect within patient 𝑖., that is,

where 𝑿" consists of all the covariates we have selected by the method of stepwise regression
above. Theoretically, a GLMM should be more reasonable than GLM, for fitting a longitudinal
dataset since it modified the estimation regarding the certain variation within individuals. For
verifying this assumption, we executed a ANOVA test between GLM and GLMM, and the
accuracy and validation of these two models were confirmed by depicting the ROC curves on
testing dataset.

## Results

#### Regression results

We initially conducted an ANOVA test between a null model (only contains intercept term)
and a variance model regarding the individual effect. The test statistic is sufficiently large
(3491.4) to reject the null hypothesis that there is no difference between these two models.
Then we could conclude that the likelihood of readmission did vary across patients. Other than
this find, we also noticed that there were variations across different level of variables, for
example, the effect of diabetes, number of inpatients, the implementation of insulin did have
distinct heterogeneous effects on the probability of readmission, as shown in Figure 3.
Furthermore, the regression result of GLM and GLMM are summarized in Table 2 and Table 3,
providing statistically significant estimations for variables: number of inpatients and length of
stays, number of diagnoses and emergency visits, as well as the status of diabetes (whether
the patient is diabetic or not). Regarding the baseline of our models was patients without being
prescribed with diabetic medication during his or her encounter, a diabetic patient definitely
has a higher risk of readmission. For instance, if a patient is diabetic, then the odds of being
readmitted within in 30 days should be exp(0.197)=1.22 multiply the odds of non-diabetic
patient. Likewise, the implementation of insulin could be regarded as an index for evaluating
the readmission risk: compared with the baseline, a patient who is prescribed with insulin, his
or her odds of readmission is higher than other medicine scenarios. Additionally, patients’
medical history and clinical care records, such as the number of inpatients and length of stay,
number of emergency he or she had visited could be also affiliated with the classification of
patients at risk, considering these records actually refer to physical quality and healthy
background.

### Model Validation 
For evaluating the validation and accuracy of GLM and GLMM regressions, we predicted the
probability of readmission on our testing dataset and depicted the ROC curves as shown in
Figure 4. The ROC curves basically display the performance of classification by plotting the
true positive rate (TPR) against the false positive rate (FPR). It indicated that the classification
by GLMM was substantially better than the one by GLM, since it had a larger value of AUC.
Therefore, GLMM model seemed to be more appropriate and efficient for fitting this dataset,
and helped us to identify the individual difference (personal effect) in the risk of readmission

```{r}
# null model 
fit0 <- glm(y ~ 1, data = train, family = binomial)
# variation in patients
fit_person <- glmer(y ~ 1 + (1 |patient_nbr), data = train, family = binomial)
anova(fit_person, fit0)

# variation in diabetes
fit_dia <- glmer(y ~  1 + (1 | diabetesMed), data = train, family = "binomial")

fit_in <- glmer(y ~  1 + (1 | number_inpatient ), data = train, family = "binomial")

fit_di <- glmer(y ~  1 + (1 | number_diagnoses ), data = train, family = "binomial")

fit_me <- glmer(y ~  1 + (1 | metformin ), data = train, family = "binomial")

fit_en <- glmer(y ~  1 + (1 | number_emergency ), data = train, family = "binomial")

fit_dis <- glmer(y ~  1 + (1 | discharge_disposition_id ), data = train, family = "binomial")

fit_leng <- glmer(y ~  1 + (1 | Length.of.Stay ), data = train, family = "binomial")

fit_medi <- glmer(y ~  1 + (1 | num_medications ), data = train, family = "binomial")

fit_ins <- glmer(y ~  1 + (1 | insulin ), data = train, family = "binomial")

dotplot(ranef(fit_dia))
dotplot(ranef(fit_in))
dotplot(ranef(fit_di))
dotplot(ranef(fit_me))
dotplot(ranef(fit_en))
dotplot(ranef(fit_dis))
dotplot(ranef(fit_leng))
dotplot(ranef(fit_medi))
dotplot(ranef(fit_ins))

# Full GLM
fit.glm <- glm(y ~ number_inpatient + diabetesMed + number_diagnoses + metformin +
                 number_emergency + discharge_disposition_id + 
                 + Length.of.Stay + num_medications + insulin,
                 data = train, family = binomial(link = "logit"))

# Full GLMM

fit.glmm <- glmer(y ~ number_inpatient + diabetesMed + number_diagnoses + metformin +
                 number_emergency + discharge_disposition_id + 
                 + Length.of.Stay + num_medications + insulin + (1 |patient_nbr),
                 data = train, family = binomial(link = "logit"))

# regression table 
(theCiMat = Pmisc::ciMat(0.95))
parTable = summary(
fit.glmm)$coef[, rownames(theCiMat)] %*% theCiMat
rownames(parTable)[1]= "Baseline"
knitr::kable(summary(fit.glmm)$coef, digits = 3)
knitr::kable(exp(parTable), digits = 3)

(theCiMat = Pmisc::ciMat(0.95))
parTable = summary(
fit.glm)$coef[, rownames(theCiMat)] %*% theCiMat
rownames(parTable)[1]= "Baseline"
knitr::kable(summary(fit.glm)$coef, digits = 3)
knitr::kable(exp(parTable), digits = 3)



# validation of model
prob1 <- predict(fit.glm, type = "response", newdata = train)
roc1 <- roc(as.factor(train$y) ~ prob1)
# The True positive rate 
TPR <- roc1$sensitivities
# The False positive rate 
FPR <- 1 - roc1$specificities

par(mfrow = c(1,2))
plot(FPR, TPR, xlim = c(0, 1), ylim = c(0, 1), type = "l", lwd = 2, col = "red", lty = 1, bty = "n", main = "GLM")
abline(0, 1, lty = 2 ,lwd = 2, col = "steelblue")
text(0.7,0.4,label = paste("AUC = ", round(auc(roc1),4)))

prob2 <- predict(fit.glmm, type = "response", train)
roc2 <- roc(as.factor(train$y) ~ prob2)

# The True positive rate 
TPR <- roc2$sensitivities
# The False positive rate 
FPR <- 1 - roc2$specificities
plot(FPR, TPR, xlim = c(0, 1), ylim = c(0, 1), type = "l", lwd = 2, col = "red", lty = 1, bty = "n", main = "GLMM")
abline(0, 1, lty = 2 , lwd = 2, col = "steelblue")
text(0.7,0.4,label = paste("AUC = ", round(auc(roc2),4)))
```

## Discussion

In this report, we focused on investigating potential determinants that could be associated
with the classification for readmitted patients at risk. We initiated our study by the process of
data cleaning and preliminary analysis: both the numerical and graphical evidence suggested
that the readmission risk varied across individuals regarding their physical quality and
healthy background. For more convinced evidence, we applied the variance component model,
GLM and GLMM regressions. The random effect of individuals was proved by ANOVA test, the
estimation results by GLM and GLMM indicated that diabetic patients who were prescribed
with insulin were suffered with a higher readmission risk than others. Additionally, the
number of inpatients and length of stays, number of emergency visits and discharge code
could be also regarded as the index for evaluating the readmission risk. The diagnosis of model
validation implicated that the GLMM could be more appreciate and accurate for classifying
patients refer to readmission.
Although several interesting facts and conclusions have been found and developed, further
work need to be considered for improving the integrity of this study. For instance, the initial
dataset had an imbalanced issue, although we divided the dataset by training and testing
dataset, the potential problem of biased estimations could not be eliminated. More efficient
and logical methods should be considered, for instance, K-fold Cross-Validation. Additionally,
we selected our variables by the method of stepwise regression on AIC criteria, which is simple
and neat, however we may consider to compare it with other methods like LASSO and Elastic
net. We also noticed that although the GLMM could be performed as an accurate classification
on readmission risk, it was extremely time-consuming by running R, hence we need to
improve our computational efficiency and accuracy in our further work.

## Reference

[1] Damian M. Predicting Diabetic Readmission Rates: Moving Beyond Hba1c. Curr Trends
Biomedical Eng & Biosci. 2017.
[2] Hempstalk, Kathryn & Mordaunt, Dylan. (2016). Improving 30-day readmission risk
predictions using machine learning. in Health Informatics New Zealand.
[3] Goudjerkan, Ti'Jay & Jayabalan, Manoj. (2019). Predicting 30-day Hospital Readmission
for Diabetes Patients Using Multilayer Perceptron. International Journal of Advanced
Computer Science and Applications. 10. 268-275.
[4] Fonarow, G. C., Konstam, M. A., & Yancy, C. W. (2017). The hospital readmission reduction
program is associated with fewer readmissions, more deaths: time to reconsider.
[5] Efroymson, M. (1966). Stepwise regression–a backward and forward look. Florham Park,
New Jersey.
[6] Myers, R. H., & Montgomery, D. C. (1997). A tutorial on generalized linear models. Journal
of Quality Technology, 29(3), 274-291.
[7] Wolfinger, R., & O'connell, M. (1993). Generalized linear mixed models a pseudolikelihood
approach. Journal of statistical Computation and Simulation, 48(3-4), 233-243.
[8] Gandrud, C. (2013). Reproducible research with R and R studio. CRC Press.
[9] Zumel, N., Mount, J., & Porzak, J. (2014). Practical data science with R (pp. 101-104).
Shelter Island, NY: Manning.

